---
title: "Cytangle: Animations"
author: "Kevin R. Coombes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: yeti
    highlight: kate
---
  
```{r setup, include=FALSE, results="hide"}
knitr::opts_chunk$set(echo = TRUE)
```
```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Overview
We are preparing CyTOF (mass cytometry) data from Greg Behbehani for analysis.
There are two parallel experiments from the same original set of samples, which
consists of bone marrow cells from patients
with acute myeloid leukemia (AML) or from normal controls. Experiment A studies
cell cycle proteins while experiment B studies B-cell receptor signaling
proteins. Both experiments use the same set of cell surface markers to identify
the individual type of white cells present in the data.

The goal of this analysis is to create animations for use in presentations of
this material. Much of the first part of the code is copied verbatim from report
`x07-threeway`.

# Getting Started
Start by defining the default paths.
```{r paths}
source("r00-paths.R")
```
We will use these packages from the R library.
```{r libs}
library(ClassDiscovery)
library(Polychrome)
p34 <- palette36.colors(36)[3:36]
library(TDA)
```
Now we load the most basic information about the experiment.
```{r basics}
load(file.path(paths$clean, "isotopes.Rda"))
load(file.path(paths$clean, "groups.Rda"))
```
Tube A contains data for the cell cycle proteins
```{r tubeA}
load(file.path(paths$clean, "cellCountsA.Rda"))
```
And we load scripts implementing a couple of reusable functions.
```{r scripts}
source("u01-justOverlay.R")
source("u02-nodeMeans.R")
```

# Setting Up the Graph
Here we get the graph for our standard SPADE plot for this data set.
```{r setupGraph}
mst <- read_graph(file.path(paths$spade, "mst.gml"), format="gml")
mst <- set_vertex_attr(mst, "size", value=3)   # shrink the nodes
mst <- set_vertex_attr(mst, "label", value="") # hide the labels
```

## Principal Components Analysis
Now we load the data for the nodes of interest. Note first that
these are the nodes that were picked in report `x06`, Second, we
are getting the cell cycle data from Tube A.

```{r ourNodes}
load(file.path(paths$scratch, "pickedNodes.Rda"))
picked <- as.list(picked)
earlyB <- paste("node", picked$early, ".Rda", sep="")
matureB <- paste("node", picked$mature, ".Rda", sep="")
samp <- "CR6"
```

Here is a principal components analysis of the data from the 
early B-cells from node `r picked$early`.
```{r spcaE}
load(file.path(paths$cluster, "A", samp, earlyB))
early <- tempMat[, Atarget]
dim(early) # 216 x 18
spcaE <- SamplePCA(t(early))
mycols <- c(early="#FF8C00", mature="blue")
```

Here is the corresponding analysis for the mature B cells from node
`r picked$mature`.
```{r spcaM}
load(file.path(paths$cluster, "A", samp, matureB))
mature <- tempMat[, Atarget]
dim(mature) # 240 x 18
spcaM <- SamplePCA(t(mature))
```

And for the joint data set.
```{r spcaME}
ME <- rbind(mature, early)
flag <- rep(c(mycols[2], mycols[1]), times=c(nrow(mature), nrow(early)))
spcaME <- SamplePCA(t(ME))
```

## TDA Results
In report `x06`, we applied TDA to nodes from the CR6 sample.
Here we load those results for the early B cells.
```{r fig.width=12, fig.height=6, fig.cap="TDA results for early B cells."}
load(file.path(paths$scratch, "Early", "CR6-Node-108-rips.Rda"))
ripsE <- rips
rm(rips)
```

We also load the TDA resulots for mature B cells.
```{r fig.width=12, fig.height=6, fig.cap="TDA results for mature B cells."}
load(file.path(paths$scratch, "Mature", "CR6-Node-120-rips.Rda"))
ripsM <- rips
rm(rips)
```

Now we compute/get the TDA (rips) results for the combined data set.
```{r combinedRips}
f <- file.path(paths$scratch, "ripsme.Rda")
if(file.exists(f)) {
  load(f)
} else {
  ripsme <- ripsDiag(usable, 1, 5, library="Dionysus", location=TRUE)
  save(ripsme, file = f)
}
rm(f)
```

# Animation
We need to load (after installing, if necessary) the `animation` R package,
```{r anime}
if (!require("animation")) {
  install.packages("animation")
  library(animation)
}
```

## Illustrating TDA
Now we are going to prepare the first animation, which illustrates the TDA process by
letting the observed data points expand. To make things simewhat clearer, we define a
function that builds the plot as a function of the charactyer expansion factor.
```{r showme}
showme <- function(CEX) {
  plot(spcaE@scores[,1], spcaE@scores[,2], col=mycols[1],
       pch=16, cex=CEX, cex.axis=1, cex.lab=1, cex.main=2,
       main="Early B Cells",
      xlab="Component 1", ylab="Component 2")
  cyc <- ripsE$cycleLocation
  rd <- ripsE[["diagram"]]
  duration <- rd[,3] - rd[,2]
  for (j in 1:length(cyc)) {
    jj <- j %% 34
    jj <- ifelse(jj==0, 34, jj)
    c19 <- cyc[[j]]
    if (prod(dim(c19)) == 0) next
    colj <- ifelse(duration[j] > 0.6, "black", "gray60")
    if (duration[j] < 0.6) next
    for(i in 1:dim(c19)[1]) {
      lines(c19[i,,1],c19[i,,2],col=colj, lwd=4)
    }
  }
}
```
To illustrate, we show one of the frames of the eventual animation.
```{r fig.width=7, fig.height=7, fig.cap = "Early B cells in PC dimension 2."}
showme(3.5)
```

Next, we set up a directory to store the animation.
```{r animeDir}
animeDir <- file.path(paths$extra, "ANIME")
if (!file.exists(animeDir)) dir.create(animeDir)
```

Now we actually create the animation. This is tricky/annoying, since the `saveHTML`
function only works in the current directory
```{r}
home <- getwd()
setwd(animeDir)
temp <- seq(1, 11, length=128)
rs <- c(temp, rev(temp))

saveHTML({for(R in 1:length(rs)) {
            showme(rs[R])
            ani.pause()
          }},
         img.name = "tda", autoplay = FALSE,
         interval = 0.04, imgdir = "ani-tda", htmlfile="01-early-tda.html",
         ani.height=600, ani.width = 600,
         title="TDA In Action",
         description="Expanding points to find loops.")
setwd(home)
```


## Illustrating TDA with pooled early and mature B cell data.

We define the function to produce plots for the pooled data.
```{r showme2}
showme2 <- function(CEX) {
  plot(spcaME@scores[,1], spcaME@scores[,2], col=flag,
       pch=16, cex=CEX, cex.axis=1, cex.lab=1, cex.main=2,
       main="B Cells",
       xlab="Component 1", ylab="Component 2") # use for grant?
  legend("bottomleft", c("Mature", "Early"), pch=16, col=c(mycols[2], mycols[1]))
  cyc <- ripsme$cycleLocation
  rd <- ripsme[["diagram"]]
  duration <- rd[,3] - rd[,2]
  for (j in 1:length(cyc)) {
    jj <- j %% 34
    jj <- ifelse(jj==0, 34, jj)
    c19 <- cyc[[j]]
    if (prod(dim(c19)) == 0) next
    colj <- ifelse(duration[j] > 0.6, "forestgreen", "gray60")
    if (duration[j] < 0.6) next
    for(i in 1:dim(c19)[1]) {
      lines(c19[i,,1],c19[i,,2],col=colj, lwd=4)
    }
  }
}
```
Here is an illustrative frame.
```{r fig.width=7, fig.height=7, fig.cap="Illustrating TDA with pooled data."}
showme2(3.5)
```

Then we create the animation
```{r reanim}
home <- getwd()
setwd(animeDir)
saveHTML({for(R in 1:length(rs)) {
            showme2(rs[R])
            ani.pause()
          }},
         img.name = "tda2", autoplay = FALSE,
         interval = 0.04, imgdir = "ani-tda2", htmlfile="02-early-mature-tda.html",
         ani.height=600, ani.width = 600,
         title="TDA In Action, Again",
         description="Expanding points to find cell cycle.")
setwd(home)
```


# Expression
In this section, we wnat to plot te expression patterns of the cell cycle data on top
of the TDA plot. Here is a function to generate one such plot.
```{r}
showme3 <- function(pal, H, CHOP = 4) {
  X <- scale(ME[, H])
  M <- max(abs(X))
  if (M > CHOP) {
    M <- CHOP
    X[X > CHOP] <- CHOP
    X[X < -CHOP] <- -CHOP
  }
  XC <- pal[cut(X, breaks = seq(-M, M, length=31), 
                labels=FALSE, include.lowest=TRUE)]
  opar <- par(c("mfrow","mai"))
  layout(matrix(1:2, nrow=1), widths = c(6, 1), heights = 1)
  on.exit(par(opar))
  plot(spcaME@scores[,1], spcaME@scores[,2], col=XC,
       pch=16, cex=1.5, cex.axis=1, cex.lab=1, cex.main=2,
       main=paste("B Cells,", H),
       xlab="Component 1", ylab="Component 2")
  cyc <- ripsme$cycleLocation
  rd <- ripsme[["diagram"]]
  duration <- rd[,3] - rd[,2]
  for (j in 1:length(cyc)) {
    jj <- j %% 34
    jj <- ifelse(jj==0, 34, jj)
    c19 <- cyc[[j]]
    if (prod(dim(c19)) == 0) next
    colj <- ifelse(duration[j] > 0.6, p34[1], "gray60")
    if (duration[j] < 0.6) next
    for(i in 1:dim(c19)[1]) {
      lines(c19[i,,1],c19[i,,2],col=colj, lwd=4)
    }
  }
  par(mai=c(1.5, 0.75, 1.5, 0.1))
  image(1, seq(-M, M, length=31), matrix(1:31, nrow=1), col = pal,
        xaxt = "n", xlab="", ylab="Standardized Units")
  par(opar)
}
```
And here is an example:
```{r fig.width=7, fig.height=6, fig.cap="CD99 expression in early and mature B cells."}
pal <- colorRampPalette(c("cyan", "gray70", "magenta"))(31)
showme3(pal, "CD99")
```

```{r echo=FALSE, eval = FALSE}
foo <- as.matrix(dist(ME))
dim(foo)
goo <- apply(foo, 1, sort)
roo <- apply(foo, 1, order)
plot(c(1, 456), c(0, 14.5), type = "n")
for (i in 1:ncol(goo)) lines(goo[,i])
abline(v = 45, col = "red")
abline(h = 4.5, col = "orange")
dang <- sapply(1:nrow(foo), function(I) {
  v <- roo[1:45, I]
  apply(ME[v,], 2, mean)
})
dong <- sapply(1:nrow(foo), function(I) {
  v <- which(foo[I,] < 4.5)
  apply(ME[v,], 2, mean)
})

# bimodals: PCNA, KI-67, Cyclin B, pRb, CD99, p56, DNA-2
opar <- par(mfrow = c(6,3), mai = c(0.2, 0.2, 0.82, 0.2))
for (I in 1:nrow(dang)) {
  hist(dang[I,], breaks = 35, main = rownames(dang)[I])
}
par(opar)


mx <- apply(ME, 1, which.max)
mn <- apply(ME, 1, which.min)

ME2 <- ME[order(mx, mn), ]
image(1:456, 1:18, ME2)
plot(apply(ME2,2, mean), type = "b")
```
```{r, echo=FALSE, eval = FALSE}
showme3(pal, "PCNA")
showme3(pal, "pRb")
showme3(pal, "pS6")
showme3(pal, "DNA-2")
```

```{r}
fdir <- file.path(paths$extra, "loopexpr")
if (!file.exists(fdir)) dir.create(fdir)
for (H in colnames(ME)) {
  png(file.path(fdir, paste(H, ".png", sep="")), width=700, height=600, bg="white")
  showme3(pal, H)
  dev.off()
}
```

# 3D Animations
We would also like to make some 3D animations. We can't do this directly in `saveHTML` from
the  `animation` package, since it doesn't know to use the function `rgl.snapshot` to save
PNG files. So, we can simlpy create all the frames, and then manually edit te appropriate
HTML and JavaScript files that go along with it. (I hate that non-reproducible step.)

Of course, we need another R package.
```{r rgl}
if (!require("rgl")) {
  install.pacakges("rgl")
  library(rgl)
}
```

We are going to start with an animation that shows moving circular sections of a sphere.
```{r circOnSphere}
foo <- diag(3)
foo
po <- ellipse3d(foo, subdivide=4, t = 1) # smooth sphere of rad1us 1

sdir <- file.path(animeDir, "circOnSphere")
if (!file.exists(sdir)) dir.create(sdir)
temp <- seq(-1, 1, length=256)
np <- 1000
theta <- seq(0, 2*pi, length = np)
for(I in 1:length(temp)) {
  x0 <- temp[I]
  x <- rep(x0, np)
  R <- sqrt(1 - x0^2)
  y <- R*cos(theta)
  z <- R*sin(theta)
  open3d(windowRect = 50 + c(0, 0, 600, 600))
  plot3d(po, axes=FALSE, alpha=0.3, col = "gray")
  points3d(x, y, z)
  rgl.snapshot(file.path(sdir, paste("cons", I, ".png", sep="")))
  rgl.close()
}
```

Now we make a hypersphere (using time as the extra dimension).
```{r, spheres}
library(colorspace)
PAL <- rainbow_hcl(128, c=90)
pal <- c(PAL, rev(PAL))
temp <- seq(1e-5, 1, length = 128)
rads <- c(temp, rev(temp))

foo <- diag(3)
po <- ellipse3d(foo, subdivide=4, t = 1) # smooth sphere of rad1us 1

sdir <- file.path(animeDir, "spheres")
if (!file.exists(sdir)) dir.create(sdir)
for(I in 1:length(rads)) {
  open3d(windowRect = 50 + c(0, 0, 600, 600))
  plot3d(po, axes=FALSE, alpha=0, col = "white")
  foo <- diag(3)*rads[I]
  eep <- ellipse3d(foo, subdivide=4, t = 1) # smooth sphere
  plot3d(eep, col = pal[I], alpha = 0.3, add = TRUE)
  rgl.snapshot(file.path(sdir, paste("cons", I, ".png", sep="")))
  rgl.close()
}
```

```{r}
open3d(windowRect = 50 + c(0, 0, 600, 600))
plot3d(po, axes=FALSE, alpha=0, col = "white")
for(I in seq(16, 128, 16)) {
  foo <- diag(3)*rads[I]
  eep <- ellipse3d(foo, subdivide=4, t = 1) # smooth sphere
  plot3d(eep, col = pal[I], alpha = (136-I)/128, add = TRUE)
}
rgl.snapshot(file.path(paths$extra, "gumball.png"))
rgl.close()
```

## Circles make a sphere
Finally, we make a video f expanding and contracting circles which serves as a time-dependent
model of a plain old sphere.

```{r circles, fig.width=6, fig.height=6, fig.cap="A circle."}
pc <- function(R, ...) {
  theta <- seq(0, 2*pi, length=600)
  x <- R*cos(theta)
  y <- R*sin(theta)
  plot(x, y, xlim=c(-2, 2), ylim=c(-2,2), ...)
}

pc(1, type="l", lwd=3)
```

```{r anicirc}
temp <- seq(0, 1.8, length=64)
rs <- c(temp, rev(temp))

home <- getwd()
setwd(animeDir)
saveHTML({for(R in 1:length(rs)) {
            pc(rs[R], type = "l", lwd = 3)
            ani.pause()
          }},
         img.name = "circ", autoplay = FALSE,
         interval = 0.04, imgdir = "circdir", htmlfile="04-circ.html",
         ani.height=600, ani.width = 600,
         title="Blip Circle",
         description="Growing and shrinking circle.")
setwd(home)
```


# Appendix
These computations were performed in the following environment:
```{r si}
sessionInfo()
```
and in the following directory:
```{r where}
getwd()
```

