---
title: "WayFindR: Computing Graph Structures on WikiPathways"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{WayFindR: Computing Graph Structures on WikiPathways}
  %\VignetteKeywords{pathways, graphs, cycles}
  %\VignetteDepends{WayFindR,igraph,XML}
  %\VignettePackage{WayFindR}
  %\VignetteEngine{knitr::rmarkdown}output: 

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Comprehending biological pathways stands as a pivotal endeavor in the realm of life sciences. However, merely possessing visual representations such as PNG or JPG files is insufficient for comprehensive analysis. The WayFindR package is a robust tool meticulously crafted for both biologists and bioinformaticians. Its primary objective is streamlining the intricate task of dissecting cellular pathways. Through seamless integration of WikiPathways (a well-known repository for biological pathways) and R's igraph package, WayFindR bestows users with the capability to transform complex pathway architectures into actionable insights.

From a mathematical perspective, control theory asserts that negative feedback is essential for effective regulation and control of dynamic systems. Negative feedback mechanisms serve as stabilizing forces, ensuring that deviations from desired states are minimized and the system returns to equilibrium. In the context of biological pathways, where intricate networks of molecular interactions govern cellular processes, the concept of negative feedback is paramount.

Biologists rely on understanding the dynamics of biological pathways to elucidate mechanisms underlying disease, identify potential drug targets, and uncover regulatory networks. Negative feedback loops within these pathways play crucial roles in maintaining cellular homeostasis, modulating responses to external stimuli, and preventing excessive or aberrant signaling.

By enabling the conversion of pathway structures into computationally tractable formats and providing tools for analysis, WayFindR facilitates the identification and characterization of negative feedback loops within pathways. This capability is invaluable for deciphering the regulatory logic governing cellular processes and designing interventions to modulate pathway activity.

In essence, by bridging the gap between biological knowledge and mathematical principles, WayFindR empowers researchers to navigate the intricate landscapes of cellular pathways with precision and insight, ultimately advancing our understanding of complex biological systems and paving the way for innovative therapeutic strategies.

# Installation
The WayFindR package is available from CRAN:

```{r}
#install.packages("WayFindR")
#install.packages("WayFindR", repos="http://R-Forge.R-project.org")
library(WayFindR)
```


# Retrieving data structure from GPML file
Let's explore the basic functionalities of WayFindR using one example file. 
Initially, we obtain a GPML file from https://www.wikipathways.org/ pertaining to a pathway that piques our interest. More specifically, we opt for the pathway labeled "Factors and pathways influencing insulin-like growth factor (IGF1)-Akt signaling (WP3850)," accessible at https://www.wikipathways.org/pathways/WP3850.html. This GPML file is integrated into a system file.

```{r}
gpmlFile <- system.file("pathways/WP3850.gpml", package = "WayFindR")
xmlfile <- XML::xmlParseDoc(gpmlFile)
```

## Edges
A user has the capability to generate a matrix with information about 
edges from the provided GPML file:

```{r}
edges <- collectEdges(xmlfile)
class(edges)
```

```{r}
dim(edges)
```
The matrix comprises 46 rows (representing edges) and 3 columns: Source, Target, and MIM (Molecular Interaction Maps). Each edge is identified by a unique ID, such as id2257fd8. These edges are directional, indicating they originate from one node and terminate at another. The Source and Target columns contain labels corresponding to these nodes. The MIM column illustrates the type of interaction occurring between the nodes.

```{r}
head(edges)
```

```{r}
tail(edges)
```


## Nodes
Likewise, users can acquire a matrix of nodes in a similar manner.

```{r}
nodes <- collectNodes(xmlfile)
class(nodes)
```

```{r}
dim(nodes)
```
In the selected pathway, there are a total of 52 nodes. The resulting matrix consists of three columns: GraphId, label, and Type. GraphId correlates to a distinct node ID extracted from the GPML file, while label corresponds to the name of a gene, protein, group, complex or biological process. The Type column contains information regarding the type of node.


```{r}
head(nodes)
```

```{r}
tail(nodes)
```
## Groups
We've already extracted all edges (Interactions) and vertices (DataNodes) from a GPML file. Another significant element within a GPML file is referred to as a "Group." However, the precise semantics of a Group are not entirely clear. In certain instances, they may denote a complex, often accompanied by a separate DataNode labeled with the attribute Type="Complex." In other cases, Groups seem to represent genes with interchangeable roles within the pathway. For instance, within the IGF1-AKT pathway, there exists an anonymous group featuring SMAD2 and SMAD3.

```{r}
groups <- collectGroups(xmlfile)
class(groups)
```

```{r}
groups$nodes
```

```{r}
groups$edges
```

MIM entry "contained" refers to an edge between a group and its constituent parts. When a group contains individual elements, such as gene products or proteins, the "contained" edge signifies the relationship between the group (container) and its components.


## Anchors
The Anchor, which is the fourth and final structure outlined in the GPML specification, serves a specific purpose. It is utilized to designate edges (or interactions, or arrows) that act as targets for other edges, instead of the typical nodes (or DataNodes, or vertices). To handle this scenario, we modify the target arrow, originally represented as A -> B, into a two-step arrow, A -> EDGE -> B. Subsequently, we transform the unconventional arrow into a simpler, more standardized representation that now points to the newly introduced EDGE node.

```{r}
links <- collectAnchors(xmlfile)
class(links)
```

```{r}
links$nodes
```

```{r}
links$edges
```

# Converting Pathways into an igraph object
The core feature of WayFindR is its ability to convert pathways from WikiPathways GPML format into igraph objects.

```{r}
library(WayFindR)
data(edgeColors)
data(edgeTypes)
data(nodeColors)
data(nodeShapes)
```

```{r}
G <- GPMLtoIgraph(xmlfile)
```

To visualize and analyze our graph, we can apply the graphopt layout algorithm from the igraph package:
```{r fig.width = 12, fig.height = 12}
set.seed(13579)
L <- igraph::layout_with_graphopt
plot(G, layout=L)
nodeLegend("topleft", G)
edgeLegend("bottomright", G)
```

# Finding cycles
Once the pathway is converted, WayFindR provides a suite of tools for in-depth analysis. Let's explore how to find cycles within the pathway graph (if they exist).

```{r}
cyc <- findCycles(G)
cyc
```

The output consists of a list of cycles extracted from a directed graph. Each cycle is represented as a sequence of vertices, where each vertex is indicated by its numerical position within the graph's vertex list. These cycles denote closed loops of vertices in the graph, signifying paths that both start and end at the same vertex.
WP3850 pathway has 5 cycles. 


Furthermore, within WayFindR, there exists a function named "interpretCycle" designed to interpret a cycle within the context of the associated graph. This function takes a cycle and the graph it belongs to as inputs. Its purpose is to extract pertinent details from the cycle, including edge IDs, arrow directions (indicating the edge direction), and labels associated with genes.

```{r}
lapply(cyc, interpretCycle, graph = G)
```
Notice that the cycles in the chosen pathway include only inhibition and stimulation processes.

Finally, we can generate a subgraph from the original graph, containing only the vertices that participate in the specified cycles. 

```{r}
S <- cycleSubgraph(G, cyc)
```


We depict the section of the graph that impacts cycles within the IGF1-AKT wikiPathway.
```{r fig.width=10, fig.height=10}
set.seed(93217)
plot(S)
nodeLegend("topleft", S)
edgeLegend("bottomright", S)

```

Here, Group1 corresponds to the the pair SMAD2, SMAD3.



