---
title: "Cytangle: Circos Plots"
author: "Kevin R. Coombes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: yeti
    highlight: kate
---
  
```{r setup, include=FALSE, results="hide"}
knitr::opts_chunk$set(fig.width=8, fig.height=5)
options(width=96)
.format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
.tag <- function(N, cap ) ifelse(.format == "html",
                                 paste("Figure", N, ":",  cap),
                                 cap)
.ntag <- function(start, caps) {
  if(.format == "html") {
    L <- length(caps)
    val <- paste("Figure ", start + (1:L), ": ", caps, sep = "")
  } else {
    val <- caps
  }
  val
}
```
```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Overview
We are preparing CyTOF (mass cytometry) data from Greg Behbehani for analysis.
There are two parallel experiments from the same original set of samples, which
consists of bone marrow cells from patients with acute myeloid leukemia (AML)
or from normal controls. Experiment A studies cell cycle proteins while
experiment B studies B-cell receptor signaling proteins. Both experiments use
the same set of cell surface markers to identify the individual type of white
cells present in the data.

The goal of this report is to examine and plot the node-sample combinations
with significant "loops" (circles) or "voids" (topologically spherical holes).

# Getting Started
## Code
Start by defining the default paths.
```{r paths}
source("r00-paths.R")
### set default paths
ripDir <- file.path(paths$scratch, "rip")
rip1Dir <- file.path(paths$scratch, "rip1d")
rip3Dir <- file.path(paths$scratch, "rip3d")
```
Load some useful packages.
```{r libs}
library(Polychrome)
p34 <- palette36.colors(36)[3:36]
suppressMessages( library(ClassDiscovery) )
suppressMessages( library(TDA) )            # to interpret rips diagrams
suppressMessages( library(circlize) )       # for circos-like plots
suppressMessages( library(ComplexHeatmap) ) # for 'Legend: and 'draw"

library(movMF)
```
Read the source code for a few utility functions.
```{r util}
suppressMessages( source("u01-justOverlay.R") ) # defines 'lay', 'ucoords', and 'overlay'
```

## Data
Now we load some generic data structures.
```{r basics}
load(file.path(paths$clean, "groups.Rda")) # defines 'groups'
summary(groups)
load(file.path(paths$clean, "isotopes.Rda")) # defines 'AMat', 'onA', 'Atarget', and 'clusmark'
dim(AMat)
head(onA)
Atarget
clusmark
```

Finally, load the collated results of the TDA.
```{r collate}
load(file.path(paths$scratch, "recollate.Rda")) # defines collate
collate$Dimension <- factor(collate$Dimension)
summary(collate)
```

# Loops
Loops or voids that we found to be interesting (because the persistence = duration is
sufficiently long) are identified by the tube (A or B), the node (a number from 1 to 483),
and the sample name. We can use that information to recover the pre-computed rips diagram.
Here is an example.
```{r example}
samname <- "CR6"
tube <- "A"
node <- 108
ripsfile <- paste("tube", tube, "-", samname, "-node", node, "-rips.Rda", sep = "")
load(file.path(rip1Dir, "A", ripsfile)) # defines 'rips' and 'keeper'
names(rips)
```

The actual data for individual cells at this node for this sample can also be loaded.
```{r nodeData}
nodefile <- paste("node", node, ".Rda", sep='')
load(file.path(paths$cluster, "A", samname, nodefile)) # defines 'tempmat'
dim(tempMat)
```

Here is a PCA plot of the target data for this tube, node, and sample )**Figure 1**).
```{r fig01, fig.width=6, fig.height=6, fig.cap = .tag(1, "Principal components plot of cells at node 108 in tube A for sample CR6.")}
M <- tempMat[, Atarget]
spcaM <- SamplePCA(t(M))
usable <- spcaM@scores[, 1:2]
plot(usable, pch = 16, col = "gray")
```

We can also extract and display the longest cycle from the TDA rips analysis. In **Figure 2**,
the points "supporting" the cycle are shown in blue, and connected by red edges. The larger
orange point marks the centoid of the cycle's support.
```{r fig02, fig.width=6, fig.height=6, fig.cap = .tag(2, "Longest 1D cycle at node 108 in tube A for sample CR6.")}
delta <- 0.5
rd <- rips[["diagram"]]
oneD <- rd[,1] == 1 # first col of rd is 'dimension'
cyc <- rips$cycleLocation[oneD]
duration <- (rd[,3] - rd[,2])[oneD]
J <- which.max(duration)
c19 <- cyc[[J]]
colj <- p34[1]
X <- as.vector(c19[,,1])
Y <- as.vector(c19[,,2])
XY <- as.data.frame(unique(cbind(X, Y)))
centroid <- apply(XY, 2, mean)

plot(usable, pch = 16, col = "gray")
for(i in 1:dim(c19)[1]) {
  lines(c19[i,,1],c19[i,,2], col=colj, lwd=2)
}
points(XY, col = p34[4], pch = 16)
points(centroid[1], centroid[2], col = p34[5], pch = 16, cex = 1.5)

```

## Angles From the Centroid
The next step is to compute the "angle" part of the position of each cell in the PCA plot
with respect to the centroid. First, here is the function that performs that computation.
```{r getAngle}
getAngles <- function(dset, centroid) {
  recentered <- sweep(dset, 2, centroid, "-") # centroid now at (0,0)
  atan2(recentered[,1], recentered[,2])
}
```

Now we can actually compute the angles (and foolishly convert them back from radians to degrees).
```{r degrees}
radians <- getAngles(usable, centroid)
degrees <- radians * 180/pi
```
We are also going to compute the average expression of each protein over sectors of width
20 degrees, centered at 15 degree intervals around the circle. Note that we have "doubled"
the period that we follow the data, which will later make it easier to fit a smooth curve
to the data without worrying about artifactual edge effects. (It also makes it easier to
deal with the special case of overlapping the position at 0 degrees.)
```{r GM}
deg <- c(degrees, degrees + 360)
od <- order(deg)
deg <- deg[od]
magic <- rbind(M, M)[od,]
GM <- t(sapply(seq(0, 345, 15), function(center) {
  lb <- center - 10
  ub <- center + 10
  set <- subset(magic, deg > lb & deg < ub)
  m.gene <- apply(set, 2, mean)
  m.gene
}))
rownames(GM) <- seq(0, 345, 15)
dim(GM)
```

## Looking at Individual Proteins
We uswe the next chunk of code to prepare a standard set of four plots for each
of the protein markers in the data set.
```{r prepGene}
# Create color list
col.ls <- c("#0000ff", "#ff0000", "#00ff00", "#ffaa00", "#ff00ff", "#00ffff")

prepGene <- function(gene, I) {
  G <- c(M[, gene], M[, gene])
  G <- G[od]
  lo <- loess(G ~ deg, span = 0.25)
  ox <- order(lo$x)

  col <- colorRamp2(c(min(G), max(G)), c("white", col.ls[2]))
  ptcol <- col(G)
  
  crackle <- (G - min(G) + 0.01)[1:(length(G)/2)]
  probs <- crackle/sum(crackle)
  set.seed(44188)
  N <- 40000
  folderol <- sample(1:length(radians), N, replace = TRUE, prob = probs)
  weighted <- degrees[folderol]
  dd <- density(weighted)
  xy <- as.matrix(usable[folderol, 1:2])
  ng <- 6
  fit <- try(lapply(1:ng, function(K) {
    movMF(xy, k = K)
  }), silent = TRUE)
  if (inherits(fit, "try-error")) fit <- NULL
  list(gene = gene, G = G, lo = lo, ox = ox, 
       ptcol = ptcol, weighted = weighted, dd = dd,
       xy = xy, fit = fit)
}

showGene <- function(object) {
  pal <- p34[c(1, 15, 4, 5, 24, 9)]
  opar <- par(mfrow = c(2,2))
  on.exit(par(opar))
  attach(object)
  on.exit(detach())
  plot(deg, G, main = gene)
  lines(lo$x[ox], lo$fitted[ox], col = "red", lwd = 2)
  lines(seq(0, 345, 15), GM[, gene], col = "purple", lwd = 2)
  abline(v = c(0, 360))
  
  plot(usable, col = ptcol, pch = 16, main = gene)
  for(i in 1:dim(c19)[1]) {
    lines(c19[i,,1],c19[i,,2], col="black", lwd=2)
  }
  points(centroid[1], centroid[2], col = "black", pch = 16, cex = 1.5)
  
  hist(weighted, breaks = 55, prob = TRUE, main = "Weighted Samples")
  lines(dd, col = "red", lwd = 2)
  
  plot(xy, col = "gray", pch = 16, main = paste("Local Peaks of", gene))
  for (I in 1:length(fit)) points(fit[[I]]$theta, pch = 16, cex = 1.5, col = pal[I])
}
```

Now we present a plethora of plots.
```{r fig03, fig.width = 9, fig.height = 9, fig.cap = .ntag(3, paste(colnames(GM), "plots."))}
if (!interactive()) {
  ignore <- sapply(colnames(GM), function(protein) {
    pg <- prepGene(protein)
    showGene(pg)
  })
}
par(mfrow = c(1,1))
```

# Circos heatmap
Here we select the six most interesting (maybe) proteins to make a circos plot of
expression around the loop-cycle.
```{r fig22, fig.width=8, fig.height=8, fig.cap = .tag(22, "Circos plot.")}
angle.df <- GM[, pal <- c("Ki-67", "PCNA", "CycB", "pRb", "CycA", "CD99")]
opar <- par(cex = 1.5, mai = c(0, 0, 0, 0))
circos.clear()
circos.par(track.height = 0.08, 
           start.degree = 90)
# For loop for each track/gene
for(i in 1:length(angle.df[1,])) {
  # If statement to add angle designations on outside of 
  # first track only
  if(i == 1) {
    data <- as.data.frame(angle.df[,i])
    col <- colorRamp2(c(min(data), max(data)), c("white", col.ls[i]))
    circos.heatmap(data, rownames.side = "outside", 
                   cluster = FALSE, col = col, rownames.cex = 1)
  } else {
    data <- as.data.frame(angle.df[,i])
    col <- colorRamp2(c(min(data), max(data)), c("white", col.ls[i]))
    circos.heatmap(data, cluster = FALSE, col = col)
  }
}
# Generate legend
lgd <- Legend(at = colnames(angle.df), title = "Genes", type = "points",
              title_position = "topleft", legend_gp = gpar(col = col.ls))
# Draw legend
draw(lgd, x = unit(0.97, "npc"), y = unit(0.05, "npc"), just = c("right", "bottom"))
par(opar)
```

# Voids
Now we extract the subset (from `collate`) that contains node-sample pairs
with significant voids. The cutoff is based on work presented in the
previous report (`x10`).
```{r siggVoid}
sigVoid <- collate[collate$Dimension == 2 & 
                     collate$Duration > 0.33,]
sigVoid$Sample <- factor(sigVoid$Sample)
sigVoid$Node <- factor(sigVoid$Node)
summary(sigVoid)
```

These auxiliary functions help get the lower-level data that we need to display some of the results.
```{r getFuncs}
get3D <- function(tube, S, N) {
 fname <- paste(paste("tube", tube, sep=''), S, N, "rips.Rda", sep='-')
  fp <- file.path(ripDir, tube, fname)
  cat(fp, "\n", file=stderr())
  load(fp)

  load(file.path(paths$clusters, tube, S, paste(N, "Rda", sep='.')))
  M <- tempMat[, get(paste(tube, "target", sep=''))]
  spca <- SamplePCA(t(M))
  usable <- spca@scores[, 1:3]
  usable
}
getCycle <- function(tube, S, N) {
  fname <- paste(paste("tube", tube, sep=''), S, N, "rips.Rda", sep='-')
  fp <- file.path(ripDir, tube, fname)
  cat(fp, "\n", file=stderr())
  load(fp)
  RD <- rips$diagram
  dure <- RD[,3] - RD[,2]
  dimn <- RD[,1]
  CY <- rips$cycleLocation
  ptr <- which(dure > 0.25 & dimn == 2)
  if (length(ptr > 1)) {
    i <- which.max(dure[ptr])
    ptr <- ptr[i]
  }
  dure[ptr]
  dimn[ptr]
  cyc <- CY[[ptr]]
  colnames(cyc) <- c("X", "Y", "Z")
  cyc
}
```

As an example, we use the first item inthe list of significant voids.
```{r}
I = 1
sigVoid[I,]
b <- sigVoid[I, "Tube"]
s <- sigVoid[I, "Sample"]
n <- sigVoid[I, "Node"]

ripsfile <- paste("tube", b, "-", s, "-", n, "-rips.Rda", sep = "")
load(file.path(ripDir, "A", ripsfile)) # defines 'rips' and 'keeper'
names(rips)

nodefile <- paste(n, ".Rda", sep='')
load(file.path(paths$cluster, "A", s, nodefile)) # defines 'tempmat'
dim(tempMat)

usable <- get3D(b, s, n)
cyc <- getCycle(b, s, n)
centroid <- apply(cyc, 2, mean)
recent <- sweep(usable, 2, centroid, "-")
```

In **Figure 23**, we show the expression levels of `pRb` as a function of spherical
coordinates where the centroid of the void support is taken as the center of an actual 
sphere.
```{r fig23, fig.width=9, fig.height=4.5, fig.cap = .tag(23, "Levels of pRb in spherical coordinates.")}
r <- sqrt(apply(recent^2, 1, sum))
phi <- atan2(recent[,2], recent[,1])
psi <- acos(recent[,3]/r)
E <- tempMat[, "pRb"]
col <- colorRamp2(c(min(E), max(E)), c("white", col.ls[1]))
ptcol <- col(E)
wicked <- loess(E ~ phi + psi, span = 0.3)
phi1 <-  seq(-pi, pi, length = 51)
psi1 <- seq(0, pi, length = 26)
daft <- data.frame(phi = rep(phi1, times = 26),
                   psi = rep(psi1, each = 51))
eek <- matrix(predict(wicked, daft), ncol = 26)
pal <- viridisLite::viridis(64)

opar <- par(mfrow =c(1,2))
plot(phi, psi, col = ptcol, pch = 16)
image(phi1, psi1, eek, col = pal)
par(opar)
```

Next, we want to unwrap the longitude coordinates twice, so we can get a smoother
fit to the line where we sliced opened the sphere (**Figure 24**).
```{r fig24, fig.width=9, fig.height=4.5, fig.cap = .tag(24, "Doubly wrapped.")}
phi2 <- c(phi, phi + 2*pi)
psi2 <- c(psi, psi)
E2 <- c(E, E)
wicked2 <- loess(E2 ~ phi2 + psi2, span = 0.2) # fits the duplicated model
phi3 <-  seq(-pi, 3*pi, length = 101) # full horizontal; coordinates
psi3 <- seq(0, pi, length = 26)      # 
daft <- data.frame(phi2 = rep(phi3, times = 26),
                   psi2 = rep(psi3, each = 101))
eek2 <- matrix(P <- predict(wicked2, daft), ncol = 26, byrow = FALSE)

opar <- par(mfrow = c(1,2))
plot(phi2, psi2, pch = 16, col = c(ptcol, ptcol))
image(phi3, psi3, eek2, col = pal) 
par(opar)
```


In **Figure 25**, we extract the middle  portion of that image.
```{r fig25, fig.width=6, fig.height=6, fig.cap = .tag(25, "Extracted center.")}
W <- phi3 >= 0 & phi3 <= 2*pi
image(phi3[W], psi3, eek2[W,], col = pal, zlim = range(eek2, na.rm = TRUE)) 
```

```{r fig26plus, fig.width=8, fig.height = 12, fig.cap = "Spherical protein expression patterns."}
exprs <- list()
opar <- par(mfrow = c(3, 2))
for (protein in colnames(GM)) {
  E <- tempMat[, protein]
  E2 <- c(E, E)
  wicked2 <- loess(E2 ~ phi2 + psi2, span = 0.2) # fits the duplicated model
  phi3 <-  seq(-pi, 3*pi, length = 101) # full horizontal; coordinates
  psi3 <- seq(0, pi, length = 26)      # 
  daft <- data.frame(phi2 = rep(phi3, times = 26),
                     psi2 = rep(psi3, each = 101))
  eek2 <- matrix(P <- predict(wicked2, daft), ncol = 26, byrow = FALSE)
  exprs[[protein]] <- eek2
  image(phi3[W], psi3, eek2[W,], col = pal, zlim = range(eek2, na.rm = TRUE), main = protein) 
}
par(opar)
```

```{r}
scaled <- lapply(exprs, function(M) {
  X <- max(M, na.rm = TRUE)
  N <- min(M, na.rm = TRUE)
  (M-N)/(X-N)
})
library(abind)
goop <- abind(scaled, along=3)
mex <- apply(goop, 1:2, which.max)
range(mex, na.rm = TRUE)
opar <- par(mfrow = c(1,2))
plot(phi2, psi2, pch = 16, col = "gray60", ylim = c(0, pi), xlim = c(0, 2*pi))
image(phi3, psi3, mex, col = pal, ylim = c(0, pi), xlim = c(0, 2*pi))
par(opar)
```

# Appendix
These computations were performed in the following environment:
```{r si}
sessionInfo()
```
and in the following directory:
```{r where}
getwd()
```

