---
title: "RPointCloud: Regulatory T Cells"
author: "Kevin R. Coombes"
data: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RPointCloud: Regulatory T Cells}
  %\VignetteKeywords{TDA, topological data analysis, regulatory T cells}
  %\VignetteDepends{RPointCloud,igraph,Polychrome,Mercator,ClassDiscovery,ape}
  %\VignettePackage{RPointCloud}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r opts, echo=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=5)
oopt <- options(width=96)
.format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
.tag <- function(N, cap ) ifelse(.format == "html",
                                 paste("Figure", N, ":",  cap),
                                 cap)
```
# Introduction
We want to illustrate the `RPointCloud` package (Version `r packageVersion("RPointCloud")`)
with a single-cell mass cytometry data set. Not surprisingly, we start by loading the package.
```{r RPointCloud}
library(RPointCloud)
```
We also load a lot of useful packages (some of which will eventually get incorporated
into the package requirements).
```{r libpack}
library(TDA)
library(Polychrome)
data(Dark24)
data(Light24)
library(Mercator)
library(igraph)
library(ClassDiscovery)
library(PCDimension)
library("ape")
```
Now we fetch the sample data set that is included with the package.
```{r CLL}
data(treg)
ls()
```

```{r echo=FALSE, eval = FALSE}
dmat <- distanceMatrix(dset, "pearson")
picked <- Mercator::downsample(target = 250, 
                               distanceMat = as.matrix(dmat),
                               cutoff = 1E-6)
treg <- dset[, picked]
tmat <- distanceMatrix(treg, "pearson")
rm(picked)
set.seed(84263)
rip <- ripsDiag(tmat, 2, 0.7, "arbitrary", "Dionysus", TRUE)
save(rip, treg, tmat, file = "treg.rda")
```

# TDA Built-in Visualizations of the Rips Diagram
Here are some plots of the `TDA` results using tools from the original package.
(I am not sure what any of these are really good for.)
```{r fig01, fig.width = 9, fig.cap = .tag(1, "The Rips barcode diagram from TDA.")}
diag <- rip[["diagram"]]
opar <- par(mfrow = c(1,2))
plot(diag, barcode = TRUE, main = "Barcode")
plot(diag, main = "Rips Diagram")
par(opar)
rm(opar)
```

```{r fig02, fig.width = 12, fig.cap = .tag(2, "Landscape, silhouette, and lambda-cluster plots of the Rips diagram, from TDA.")}
L <- TDA::landscape(diag, KK = 1)
S <- TDA::silhouette(diag)
crt <- TDA::clusterTree(as.matrix(tmat), k = 5, dist = "arbitrary")

opar <- par(mfrow = c(1, 3))
plot(L, type = "l", main = "Landscape")
plot(S, type = "l", main = "Silhouette")
plot(crt, type = "lambda",main = "Lambda Cluster Tree")
par(opar)
rm(L, S, opar)
```

# Mercator Visualizations of the Underlying Data and Distance Matrix
```{r Mercator}
M <- Mercator(tmat, metric ="pearson", method = "mds", K = 8)
M <- addVisualization(M, "hclust")
M <- addVisualization(M, "tsne")
M <- addVisualization(M, "umap")
M <- addVisualization(M, "som")
M@palette <- Light24
set.seed(72345)
clue <- kmeans(t(treg), centers = 8, iter.max = 100, nstart = 20)
M <- setClusters(M, clue$cluster)
```

```{r fig03, fig.width = 9, fig.height = 12, fig.cap = .tag(3, "Mercator Visualizations of the distance matrix.")}
opar <- par(mfrow = c(3,2), cex = 1.1)
plot(M, view = "hclust")
plot(M, view = "mds", main = "Mult-Dimensional Scaling")
plot(M, view = "tsne", main = "t-SNE")
plot(M, view = "umap", main = "UMAP")
barplot(M, main = "Silhouette Width")
plot(M, view = "som", main = "Self-Organizing Maps")
par(opar)
rm(opar)
```

# Dimension Zero
Here is a picture of the "zero-cycle" data, which can also be used ultimately to cluster
the points (where each point is a patient). The connected lines are similar to a
single-linkage clustering structure, showing when individual points are merged together
as the TDA parameter increases.
```{r fig04, fig.cap = .tag(4, "Hierarchical connections between zero cycles.")}
nzero <- sum(diag[, "dimension"] == 0)
cycles <- rip[["cycleLocation"]][2:nzero]
W <- M@view[["umap"]]$layout
plot(W, main = "Connected Zero Cycles")
for (cyc in cycles) {
  points(W[cyc[1], , drop = FALSE], pch = 16,col = "red")
  X <- c(W[cyc[1], 1], W[cyc[2],1])
  Y <- c(W[cyc[1], 2], W[cyc[2],2])
  lines(X, Y)
}
```

# Using iGraph
We can convert the 0-dimensional cycle structure into a dendrogram, by first passing them
through the `igraph` package. We start by putting all the zero-cycle data together, which
can be viewed as an "edge-list" from the `igraph` perspective.
```{r igraph}
edges <- t(do.call(cbind, cycles)) # this creates an "edgelist"
G <- graph_from_edgelist(edges)
G <- set_vertex_attr(G, "label", value = attr(tmat, "Labels"))
```
Note that we attached the sample names to the graph, obtaining them from the daisy
distance matrix. Now we use two different algorithms to decide how to layout the graph.
```{r layouts}
set.seed(2734)
Lt <- layout_as_tree(G)
L <- layout_with_fr(G)
```


```{r fig05, fig.cap = .tag(5, "Two igraph depictions of the zero cycle structure."), fig.width = 10}
opar <- par(mfrow = c(1,2), mai = c(0.01, 0.01, 1.02, 0.01))
plot(G, layout = Lt, main = "As Tree")
plot(G, layout = L, main = "Fruchterman-Reingold")
par(opar)
```
Note that the Fruchterman-Reingold layout gives the most informative structure.

## Community Structure
There are a variety of community-finding algorithms that we can apply. (Communities
in graph theory are similar to clusters in other machine learning areas of study.)
"Edge-betweenness" seems  to work best.
```{r keg}
keg <- cluster_edge_betweenness(G) # 19
table(membership(keg)) 
pal <- Light24[membership(keg)]
```

The first line in the next code chunk shows that we did actually produce a tree.
We explore three different ways ro visualize it
```{r fig06, fig.width = 6, fig.height = 6, fig.cap = .tag(6, "Community structure, simplified.")}
is.hierarchical(keg)
H <- as.hclust(keg)
H$labels <- vertex_attr(G, "names")
K <-  12
colset <- Light24[cutree(H, k=K)]
G2 <- set_vertex_attr(G, "color", value = colset)
e <- 0.01
opar <- par(mai = c(e, e, e, e))
plot(G2, layout = L)
par(opar)
```

```{r fig08, fig.width=7, fig.height = 7, fig.cap = .tag(8, "Cladogram realization, from the ape package.")}
P <- as.phylo(H)
opar <- par(mai = c(0.01, 0.01, 1.0, 0.01))
plot(P, type = "u", tip.color = colset, cex = 0.8, main = "Ape/Cladogram")
par(opar)
rm(opar)
```


## Visualizing Features
In any of the "scatter plot views" (e.g., MDS, UMAP, t-SNE) from Mercator, we may want to
overlay different colors to represent different clinical features.
```{r views}
U <- M@view[["mds"]]
V <- M@view[["tsne"]]$Y
W <- M@view[["umap"]]$layout
```

```{r fig10, fig.width = 9, fig.cap = .tag(10, "UMAP visualizations with clinical features.")}
FOXP3 <- Feature(treg["FOXP3",], "FOXP3", c("pink", "skyblue"), c("Low", "High"))
CTLA4 <- Feature(treg["CTLA4", ], "CTLA4", c("green", "magenta"), c("Low", "High"))
opar <- par(mfrow = c(1,2))
plot(W, main = "UMAP; FOXP3", xlab = "U1", ylab = "U2")
points(FOXP3, W, pch = 16, cex = 1.4)
plot(W, main = "UMAP; CTLA4", xlab = "U1", ylab = "U2")
points(CTLA4, W, pch = 16, cex = 1.4)
par(opar)
rm(opar)
```


```{r cleanup}
options(oopt)
#rm(list = ls())
```
